What We Eval in the Shadows

### 2021-04-14 analysis/side-effects.Rmd
Side effects

@@

```{r}
SE_all_N <- sum(SE_all$N)
SE_all_sites <- count(SE_all, eval_call_srcref)
SE_all_calls <- count(SE_all, cid)

@@

SE_N <- sum(SE$N)
SE_calls <- count(SE, cid)
SE_sites <- count(SE, eval_call_srcref)
SE_missing_N <- sum(SE_missing_sites$N)

@@

  r("User sites", SE_sites),
  r("User sites ratio", ratio(SE_sites, SE_all_sites))
)
```

@@

```{r functions}
SE <- mutate(
  SE, 
  class=case_when(
    source == str_c("call:", caller_function)           ~ "Local",
    str_starts(source, fixed("call:"))                  ~ "Function",
    str_detect(source, "^R6:.*R6.*") | source=="???S4"  ~ "Object",
    source == "pkg:global"                              ~ "Global",
    str_starts(source, fixed("pkg:"))                   ~ "Package",
    TRUE                                                ~ "Synthetic"
  ),
  caller=str_replace(eval_call_srcref, "::(.*)::(.*)::.*", "\\1::\\2")
)
```

@@

```{r}
top_sites_classes <- cs_count(sites_classes, classes) %>% filter(cp<=91)
```

```{r}
print(
  xtable(
    mutate(
      top_sites_classes,
      p=fmt(p, suffix="%", digit=1),
      cp=fmt(cp, suffix="%", digit=1),
    ), 
    align="llrrr",
  ),
  file=file.path(TAGS_DIR, "table-top-sites-classes-se.tex"),
  comment=FALSE, 
  include.colnames=FALSE,
  include.rownames=FALSE,
  only.contents=TRUE,
  timestamp=FALSE
)
```

@@

```{r}
single_class_funs <- 
  left_join(
    mutate(ES, fun=str_c(package, fun_name)), 
    sites_classes, 
    by="eval_call_srcref"
  ) %>% 
  group_by(fun) %>% 
  filter(all(!is.na(n_class))) %>%
  summarize(n_classes=n_distinct(classes), classes=str_c(unique(classes), collapse=";"))

cs_count(single_class_funs, n_classes)
cs_count(single_class_funs, classes)
```

```{r}
single_class_funs %>%
  mutate(
    classes=if_else(str_detect(classes, "[,;]"), "\\emph{multiple}", classes)
  ) %>%
  cs_count(classes) %>%
  mutate(
    p=fmt(p, suffix="\\%", digit=1),
    o=if_else(str_starts(classes, fixed("\\")), 0L, n)
  ) %>% 
  arrange(desc(o)) %>%
  select(-cp, -o) %>%
  xtable(
    align="llrr",
  ) %>% 
  print(
    file=file.path(TAGS_DIR, "table-top-sites-classes-funs.tex"),
    comment=FALSE, 
    include.colnames=FALSE,
    include.rownames=FALSE,
    only.contents=TRUE,
    sanitize.text.function=identity,
    timestamp=FALSE
  )
```

```{r}
cs_count(SE, eval_call_srcref, wt=N)
cs_count(SE, variable, wt=N)
```

```{r}
overview_table(
  r("Number of side effects", sum(SE$N))
)
```

>>>>

TODO: fix the srcrefs and caller packages / functions

```{r}
SE <- mutate(
  SE,
  caller=str_replace(eval_call_srcref, "::(.*)::(.*)::.*", "\\1::\\2")
)
```

@@

```{r}
S_call <- sum(S$nb_ev_calls)
S_site <- count(S, eval_call_srcref)
S_package <- count(S, eval_source)
S_funs <- sum(PKG$funs_with_eval)

SE_all_N <- sum(SE_all$N)
SE_all_sites <- count(SE_all, eval_call_srcref)
SE_all_calls <- count(SE_all, cid)

@@

SE_N <- sum(SE$N)
SE_calls <- count(SE, cid)
SE_sites <- count(SE, eval_call_srcref)
SE_funs <- count(SE, caller)
SE_pkgs <- count(SE, caller_package)
SE_missing_N <- sum(SE_missing_sites$N)

@@

  r("User sites", SE_sites),
  r("User sites ratio", ratio(SE_sites, SE_all_sites)),

  r("User functions", count(SE, caller)),
  r("User packages", count(SE, caller_package)),
)
```

@@

```{r functions}
SE <- mutate(
  SE, 
  class=case_when(
    source == str_c("call:", caller_function)           ~ "Local",
    str_starts(source, fixed("call:"))                  ~ "Function",
    str_detect(source, "^R6:.*R6.*") | source=="???S4"  ~ "Object",
    source == "pkg:global"                              ~ "Global",
    str_starts(source, fixed("pkg:"))                   ~ "Package",
    TRUE                                                ~ "Synthetic"
  )
)
```

@@

```{r}
sum_sites_classes <- 
  sites_classes %>%
  mutate(
    classes=if_else(str_detect(classes, "[,;]"), "multiple", classes)
  ) %>%
  cs_count(classes) %>%
  mutate(
    p=fmt(p, suffix="\\%", digit=1),
    o=if_else(str_starts(classes, fixed("\\")), 0L, n)
  ) %>% 
  arrange(desc(o)) %>%
  select(-cp, -o)

sum_sites_classes %>%
  mutate(classes=if_else(classes=="multiple", "\\emph{multiple}", classes)) %>%
  xtable(
    align="llrr",
  ) %>%
  print(
    file=file.path(TAGS_DIR, "table-top-sites-classes-se.tex"),
    comment=FALSE, 
    include.colnames=FALSE,
    include.rownames=FALSE,
    only.contents=TRUE,
    sanitize.text.function=identity,
    timestamp=FALSE
  )
```

```{r}
sites_class_n <- cs_count(sites_classes, n_class)

overview_table(
  r("Sites in one class", percent(filter(sites_class_n, n_class==1)$p/100))
)
```

@@

```{r}
funs_srcrefs <- 
  left_join(
    mutate(ES, fun=str_c(package, "::", fun_name)), 
    transmute(count(S, eval_call_srcref), eval_call_srcref, hit=1), 
    by="eval_call_srcref"
  ) %>%
  mutate(hit=replace_na(hit, 0)) %>%
  select(-args, -call_fun_name) %>%
  left_join(
    select(sites_classes, eval_call_srcref, classes),
    by="eval_call_srcref"
  )

funs_cov <-
  funs_srcrefs %>% 
  group_by(fun) %>%
  summarize(cov=sum(hit)/n())

funs_classes <- 
#  semi_join(funs_srcrefs, filter(funs_cov, cov==1), by="fun") %>%
  semi_join(funs_srcrefs, funs_cov, by="fun") %>%
  group_by(fun) %>%
  summarize(
    cov=sum(hit)/n(),
    classes=str_c(unique(classes[!is.na(classes)]), collapse=";")
  )

cs_count(funs_classes, cov)
cs_count(funs_classes, classes)
```

```{r}
sum_funs_classes <-
funs_classes %>%
  filter(cov==1, classes != "") %>%
  mutate(
    classes=if_else(str_detect(classes, "[,;]"), "multiple", classes)
  ) %>%
  cs_count(classes) %>%
  mutate(
    p=fmt(p, suffix="\\%", digit=1),
    o=if_else(str_starts(classes, fixed("\\")), 0L, n)
  ) %>% 
  arrange(desc(o)) %>%
  select(-cp, -o)

sum_funs_classes %>%
  mutate(classes=if_else(classes=="multiple", "\\emph{multiple}", classes)) %>%
  xtable(
    align="llrr",
  ) %>% 
  print(
    file=file.path(TAGS_DIR, "table-top-sites-classes-funs.tex"),
    comment=FALSE, 
    include.colnames=FALSE,
    include.rownames=FALSE,
    only.contents=TRUE,
    sanitize.text.function=identity,
    timestamp=FALSE
  )
```

```{r}
left_join(
  rename(sum_sites_classes, site_n=n, site_p=p),
  rename(sum_funs_classes, fun_n=n, fun_p=p)
) %>%
  mutate(
    o=if_else(classes=="multiple", 0L, site_n),
    classes=if_else(classes=="multiple", "\\emph{multiple}", classes),
    fun_n=replace_na(fun_n, "0"),
    fun_p=replace_na(fun_p, "0\\%"),
  ) %>%
  arrange(desc(o)) %>%
  select(-o) %>%
  xtable(
    align="llrrrr"
  ) %>%
  print(
    file=file.path(TAGS_DIR, "table-se-target-envs.tex"),
    comment=FALSE, 
    include.colnames=FALSE,
    include.rownames=FALSE,
    only.contents=TRUE,
    sanitize.text.function=identity,
    timestamp=FALSE
  )
```

```{r}
event_calls <- cs_count(SE, event, wt=N)
event_site <- cs_count(SE, eval_call_srcref, event) %>% cs_count(event)

left_join(
  select(event_calls, event, n_call=n, p_call=p),
  select(event_site, event, n_site=n, p_site=p),
  by="event"
) %>%
  mutate_if(is.double, ~fmt(., digit=1)) %>%
  mutate_if(is.integer, ~fmt(.)) %>%
  xtable(
    align="llrrrr"
  ) %>%
  print(
    file=file.path(TAGS_DIR, "table-se-types.tex"),
    comment=FALSE, 
    include.colnames=FALSE,
    include.rownames=FALSE,
    only.contents=TRUE,
    sanitize.text.function=identity,
    timestamp=FALSE
  )
```

```{r}
filter(SE, event=="rvl") %>% cs_count(class)
```

### 2021-04-14 paper/main.tex
typo
Side effects

\begin{table}[h]
  \centering\small\hspace{-.5cm}
\begin{minipage}{3.7cm}
  \begin{tabular}{@{}r|r|r@{}}\hline
 Kind & \#sites & \%sites \\\hline
 Function & \packageNbFunctionEnvSites &  \packageNbFunctionEnvSitePercent\\
 Synthetic & \packageNbSyntheticEnvSites & \packageNbSyntheticEnvSitePercent \\
 Global &  \packageNbStrictGlobalEnvSites & \packageNbStrictGlobalEnvSitePercent \\
 Package & \packageNbPackageNamespaceEnvSites & \packageNbPackageNamespaceEnvSitePercent \\\hline
\end{tabular}
\caption{Kinds per site} \label{tab:highlevelenvironments}
\end{minipage}\hspace{-.2cm}
\begin{minipage}{3.7cm}\centering
\begin{tabular}{@{}r|r|r@{}}\hline
 Offset & \#sites & \%sites \\\hline
  \packageCallerEnvHierarchyNamea & \packageCallerEnvHierarchySitesaRnd & \packageCallerEnvHierarchySitePercenta \\
  \packageCallerEnvHierarchyNameb& \packageCallerEnvHierarchySitesbRnd & \packageCallerEnvHierarchySitePercentb \\
  \packageCallerEnvHierarchyNamec& \packageCallerEnvHierarchySitescRnd & \packageCallerEnvHierarchySitePercentc  \\
  \ge 3& \packageNbFarAwayCallerSites &  \packageNbFarAwayCallerSitePercent \\\hline
 \end{tabular}
\caption{Function offset}\label{tab:funoffset}
\end{minipage}\hspace{-.2cm}
\begin{minipage}{3.7cm}
\begin{tabular}{@{}r|r|r@{}} \hline
Parent & \#sites & \%sites \\\hline
Function & \packageNewEnvCategorySitesa & \packageNewEnvCategorySitePercenta \\
Package & \packageNewEnvCategorySitesb &  \packageNewEnvCategorySitePercentb\\
Global & \packageNewEnvCategorySitesc & \packageNewEnvCategorySitePercentc \\
Empty & \packageNewEnvCategorySitesd & \packageNewEnvCategorySitePercentd \\\hline
\end{tabular} 
\caption{Wrapper envs.} \label{tab:newenvs}
\end{minipage}\hspace{-.2cm}
\begin{minipage}{3.7cm}\centering
 \begin{tabular}{@{}c|c|c@{}} \hline
 \#kinds & \#sites &  \%sites \\ \hline
 \packageNbCategoryEnvira & \packageNbCategoryEnvirSitesaRnd &  \packageNbCategoryEnvirPercenta\\
 \packageNbCategoryEnvirb &  \packageNbCategoryEnvirSitesbRnd & \packageNbCategoryEnvirPercentb \\
 \packageNbCategoryEnvirc & \packageNbCategoryEnvirSitescRnd &  \packageNbCategoryEnvirPercentc\\
 \packageNbCategoryEnvird & \packageNbCategoryEnvirSitesdRnd & \packageNbCategoryEnvirPercentd\\\hline
\end{tabular}\caption{Multiplicities}\label{tab:polyenvir}
\end{minipage}\hspace{-1cm}
\end{table}

@@

From the \packageNbrunsRnd programs we capture \SEAll side effects from
\SEAllCalls \eval calls in \SEAllSites sites. The challenge is again to remove
the accidental side-effects that are caused by the R virtual machine
implementation and are not related to the user code. For example, the
\c{.Random.seed} variable, which contains the state of the random number
generator is saved and restored from and to the global environment every time
user calls one of the base routines for getting random numbers. Removing them
leaves us with \SEUser side effects (\SEUserRatio) from \SEUserCalls \eval
calls (\SEUserCallsRatio) in \SEUserSites sites (\SEUserSitesRatio). Half of
the side-effects comes from three functions: \c{plyr::allocate\_column}
(allocates space for a new data frame column), \c{withr::execute\_handlers}
(executes deferred expressions) and \c{foreach::doSEQ} (executes an expression
on each element in a collection, possibly in parallel). \SEFunsNighty functions
are responsible for 90\% of side effects. The remaining is spread in \SEFunsTen
functions.

Most of the \eval sites (\SESitesInEnvirRatio) do side effects in the
environment specified by the \c{envir} parameter (\cf
Section~\ref{sec:eval-in-r}). \SESitesNotInEnvirRatio modifies other
environment and finally \SESitesBothEnvirRatio does both.
Table~\ref{tab:side-effects-targets} shows the class of the environments where
\eval side effects happen. It follows the classification described in
Section~\ref{sec:env} with two extra classes: \emph{Object} which represents a
side-effect that happens in S4 or R6 object environment, and \emph{Local} that
corresponds to a side effect in local variables of the \eval caller.

\begin{table}[h]\centering\small
  \begin{tabular}{l|r|r|r}\hline
    Environments & \#sites & \%sites & cum. \%sites \\%
    \expandableinput tag/table-top-sites-classes-se.tex
  \end{tabular}
  \caption{Target environments for \eval side-effects} \label{tab:side-effects-targets}
\end{table}

The good news is that that most \eval sites do all side effects consistently in
the same environment and that almost half of the side effects happen in either
\emph{Local} or \emph{Object} environments. These are the cases the compiler
could possibly deal with.
\todo{Finish text for Table~\ref{tab:se-funs}}

\begin{table}[h]\centering\small
  \begin{tabular}{l|r|r}\hline
    Environment & \#functions & \%functions \\%
    \expandableinput tag/table-top-sites-classes-funs.tex
  \end{tabular}
  \caption{Number of functions in which all \eval side effects happen in the same environment class} \label{tab:se-funs}
\end{table}

>>>>

\begin{table}[h]
  \centering\small\hspace{-.5cm}
\begin{minipage}{3.7cm}
  \begin{tabular}{@{}r|r|r@{}}\hline
 Kind & \#sites & \%sites \\\hline
 Function & \packageNbFunctionEnvSites &  \packageNbFunctionEnvSitePercent\\
 Synthetic & \packageNbSyntheticEnvSites & \packageNbSyntheticEnvSitePercent \\
 Global &  \packageNbStrictGlobalEnvSites & \packageNbStrictGlobalEnvSitePercent \\
 Package & \packageNbPackageNamespaceEnvSites & \packageNbPackageNamespaceEnvSitePercent \\\hline
\end{tabular}
\caption{Kinds per site} \label{tab:highlevelenvironments}
\end{minipage}\hspace{-.2cm}
\begin{minipage}{3.7cm}\centering
\begin{tabular}{@{}r|r|r@{}}\hline
 Offset & \#sites & \%sites \\\hline
  \packageCallerEnvHierarchyNamea & \packageCallerEnvHierarchySitesaRnd & \packageCallerEnvHierarchySitePercenta \\
  \packageCallerEnvHierarchyNameb& \packageCallerEnvHierarchySitesbRnd & \packageCallerEnvHierarchySitePercentb \\
  \packageCallerEnvHierarchyNamec& \packageCallerEnvHierarchySitescRnd & \packageCallerEnvHierarchySitePercentc  \\
$\ge 3$& \packageNbFarAwayCallerSites &  \packageNbFarAwayCallerSitePercent \\\hline
 \end{tabular}
\caption{Function offset}\label{tab:funoffset}
\end{minipage}\hspace{-.2cm}
\begin{minipage}{3.7cm}
\begin{tabular}{@{}r|r|r@{}} \hline
Parent & \#sites & \%sites \\\hline
Function & \packageNewEnvCategorySitesa & \packageNewEnvCategorySitePercenta \\
Package & \packageNewEnvCategorySitesb &  \packageNewEnvCategorySitePercentb\\
Global & \packageNewEnvCategorySitesc & \packageNewEnvCategorySitePercentc \\
Empty & \packageNewEnvCategorySitesd & \packageNewEnvCategorySitePercentd \\\hline
\end{tabular}
\caption{Wrapper envs.} \label{tab:newenvs}
\end{minipage}\hspace{-.2cm}
\begin{minipage}{3.7cm}\centering
 \begin{tabular}{@{}c|c|c@{}} \hline
 \#kinds & \#sites &  \%sites \\ \hline
 \packageNbCategoryEnvira & \packageNbCategoryEnvirSitesaRnd &  \packageNbCategoryEnvirPercenta\\
 \packageNbCategoryEnvirb &  \packageNbCategoryEnvirSitesbRnd & \packageNbCategoryEnvirPercentb \\
 \packageNbCategoryEnvirc & \packageNbCategoryEnvirSitescRnd &  \packageNbCategoryEnvirPercentc\\
 \packageNbCategoryEnvird & \packageNbCategoryEnvirSitesdRnd & \packageNbCategoryEnvirPercentd\\\hline
\end{tabular}\caption{Multiplicities}\label{tab:polyenvir}
\end{minipage}\hspace{-1cm}
\end{table}

@@

From the \packageNbrunsRnd programs we capture \SEAll side effects from
\SEAllCalls \eval calls in \SEAllSites sites. The challenge is again to remove
the accidental side-effects that are caused by the R virtual machine
implementation and are not related to the user code. For example, the
\c{.Random.seed} variable, which contains the state of the random number
generator is saved and restored from and to the global environment every time
user calls one of the base routines for getting random numbers. Removing them
leaves us with \SEUser side effects (\SEUserRatio) from \SEUserCalls \eval
calls (\SEUserCallsRatio) in \SEUserSites sites (\SEUserSitesRatio). These
sites are part of \SEUserFunctions in \SEUserPackages packages. \SEFunsNighty
functions are responsible for 90\% of side effects. Half of the side-effects
comes just from three functions: \c{plyr::allocate\_column} (allocates space
for a new data frame column), \c{withr::execute\_handlers} (executes deferred
expressions) and \c{foreach::doSEQ} (executes an expression on each element in
a collection, possibly in parallel).

Most of the \eval sites (\SESitesInEnvirRatio) do side effects in the
environment specified by the \c{envir} parameter (\cf
Section~\ref{sec:eval-in-r}), \SESitesNotInEnvirRatio modifies other
environment and finally \SESitesBothEnvirRatio does both.
Table~\ref{tab:se-env} shows the class of the environment where \eval side
effects happen (\emph{target environment}). It follows the classification
described in Section~\ref{sec:env} with two extra classes: \emph{Object} which
represents a side-effect that happens in S4 or R6 object environment, and
\emph{Local} that corresponds to a side effect in local variables of the \eval
caller. The table shows data for both \eval sites and functions. For a function
to have a given target environment class means that all of its \eval sites that
do side effect must have the same target environment class. Majority of the
\eval sites (\SESitesInOneClass) do all side effects consistently in one
environment class. The same happens on the function level. Almost half of the
sites and over a third of the functions do side effects in either \emph{Local}
or \emph{Object} environments. This gives a ray of hope for the compiler. Even
though it is possible to do anything anywhere, the data suggest that most
side-effects are sane.

\begin{table}[h]
  \small
  \centering
  \begin{tabular}{l|r|r|r|r}\hline
    Environments & \#sites & \%sites & \#funs. & \%funs. \\%
    \expandableinput tag/table-se-target-envs.tex
  \end{tabular}
  \caption{Most used target environments for \eval side-effects} \label{tab:se-env}
\end{table}

Table~\ref{tab:se-types} shows the proportion of the different side effects
that were recorded. In terms of calls we see mostly assignments and in terms of
sites definitions. This is expected. A subsequent \eval call will turn a
definition into an update. Variable removal, while rare, it does happen.
However, the vast majority comes from the already mentioned
\c{withr::execute\_handlers} function used to defer evaluation of an expression
to after the function exit. In our dataset, it is used almost exclusively by
the \c{tidyselect} package that uses it to remove the reference to current
quosure environment while interpreting a data frame column
selectors.\footnote{\cf \url{https://tidyselect.r-lib.org/}}

\begin{table}[h]
  \small
  \centering
  \begin{tabular}{l|r|r|r|r|r|r}\hline
    Side effect & \#events & \%events & \#calls & \%calls & \#sites & \%sites \\%
    \expandableinput tag/table-se-types.tex
  \end{tabular}
  \caption{Types of \eval side-effects} \label{tab:se-types}
\end{table}
